Supplementary information: Variational theory of non-relativistic quantum electrodynamics

### 2018-10-31 supp2.tex
about 80% there...

\documentclass[aps,prb,onecolumn,
	groupedaddress,superscriptaddress,
	amsfonts,amssymb,amsmath,floatfix,
	citeautoscript]{revtex4-1}

@@

\title{Supplementary information: Variational theory of the ground state of quantum electrodynamics}

@@

\maketitle
In this supplement, we derive a self-consistent set of equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included. We derive them in the case of a multi-electron system with electron-electron interactions in three spatial dimensions with an arbitrary number of photonic modes. The resulting equations are very similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field theory orbitals. The one-dimensional problem studied for concreteness in the main text is a special case of these equations for a single-electron emitter in a finite-dimensional space and in a one-dimensional cavity.
\section{Equations for the ground state of the coupled light-matter system}

The QED Hamiltonian takes the form:
\begin{equation}
H = H_{mat}+H_{em}+H_{int},
\end{equation}
where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:

%\begin{align}
%H_{el} = &\int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
%\end{align}
\begin{equation}
H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}\Jadd{(x)} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{equation}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. \Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}

Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$. This renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.

The interaction Hamiltonian takes the form:
%\begin{align}
%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
%\end{align}
\begin{equation}
H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\Jadd{(x)}\cdot\nabla +  \nabla \cdot \mathbf{A}\Jadd{(x)})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\Jadd{(x)}\psi\Jadd{(x)}\mathbf{A}^2(x).
\end{equation}

This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators. The electron field operator takes the form:
\begin{equation}
\psi(x) = \sum_n \psi_n(x)c_n.
\end{equation}
In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form
\begin{equation}
\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right).
\end{equation}
In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies. We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$

With the Hamiltonian set up, we now move to describe a variational theory of the ground state. In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.

We choose as our ground state
\begin{equation}
|\Omega\rangle = |\psi\rangle \otimes |\phi\rangle,
\end{equation}
where
\begin{equation}
|\psi\rangle = \prod\limits_n c_n^{\dagger}|0_n\rangle,
\end{equation}
and where
\begin{equation}
|\phi\rangle = \bigotimes_i|0_i\rangle.
\end{equation}
In such an ansatz $|\psi\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons. Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. Intuitively, this happens because the electromagnetic field is described by an electric and magnetic field, which are related to $\mathbf{F}_i$ and $\partial_t \mathbf{F}_i \rightarrow \omega_i\mathbf{F}_i$.

The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right).
\end{equation}
\end{widetext}
\Jadd{(say where the $\lambda$ is coming from)}

\subsection{System of equations for the ground state}
We now proceed with the program of evaluating the expectation value and carrying out the functional minimization with respect to all of the free parameters.

The minimization with respect to the electron orbital $\psi_i^*$ \Jadd{($\delta\mathcal{L}/\delta\psi_i^*=0$) }yields the equation:
% \begin{widetext}
% \begin{equation}
% \left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) + \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
% \end{equation}
% \end{widetext}
\begin{align}
\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n\Jadd{(x)}|^2\psi_i(x)  = E_i\psi_i(x).
\end{align}
This differs from the usual Hartree-Fock equation of electronic theory by the last term on the left-hand-side, representing a polarization induced by the $\mathbf{A}^2$ term.

The minimization with respect to the photonic mode function $\mathbf{F}_n^*$ gives:
\begin{equation}
\left(\epsilon\omega_n+\frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega_n} \right)\mathbf{F}_n + \frac{c^2}{\omega_n}\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = 2\lambda_n\omega_n\mathbf{F}_n.
\end{equation}
Meanwhile, the minimization with respect to frequency $\omega_n$ gives:
\begin{equation}
\int d^3x\left(\epsilon - \frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega^2_n}\right)|\mathbf{F}_n|^2 - \frac{c^2}{\omega_n^2}\mathbf{F}_n^*\cdot\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n=0.
\end{equation}
Equations (12) and (13), to be mutually consistent, force $\lambda_n = \omega_n$, reducing Equation (12) to the equivalent form:
\begin{equation}
\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = \left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n,
\end{equation}
where $\omega_p \equiv \frac{e^2}{2m\epsilon_0}\sum_n|\psi_n(x)|^2$ is the position-dependent plasma frequency. Notice that taking the divergence of both sides of Equation (14) requires that
\begin{equation}
\nabla\cdot\left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n = 0,
\end{equation}
which is a statement of Gauss' law for the electric field.

\section{Lamb shift correction to the equations for the quantum electrodynamical ground state}

>>>>

\documentclass[aps,prb,onecolumn,preprint,
	groupedaddress,superscriptaddress,
	amsfonts,amssymb,amsmath,floatfix,
	citeautoscript]{revtex4-1}

@@

\title{Supplementary information: Variational theory of non-relativistic quantum electrodynamics}

@@

\maketitle
In this supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included. We derive them in the case of a multi-electron system with electron-electron interactions in three spatial dimensions with an arbitrary number of photonic modes. The resulting equations are very similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field theory orbitals. We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Figure (2) of the main test.

\section{Self-consistent lamb shift correction to the equations for the quantum electrodynamical ground state}
